Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-07 paper/paper.tex
new plots
Correctness proof

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \{\psi_{i}\}, \sigma')\!:\!S, H, \kappa, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e, \sigma'$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (\CONS~x~y), \sigma$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma'$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma' )\!:\!S, H, (\CAR~x), \sigma$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma'$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x), \{\psi_{i}\}, \sigma)\!:\!S, H, s,\bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x} $}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \{\psi_{i}\}, \sigma')\!:\!S, H, (+~x~y), \sigma$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma'$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{x}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y), \sigma$ \\ $\rightsquigarrow \rho', (\rho, y, (+~x~y), \{\psi_{i}\}, \sigma)\!:\!S, H, s, \bigcup_{\psi \in \{\psi_{i}\}}\mmap{}{\psi}{y} $}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x}), \sigma$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}, \sigma$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\pi:\SIF~\psi:x~e_1~e_2), \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \pmap{}(\pi), \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma')\!:\!S, H, (\SRETURN~x), \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, \sigma'$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\psi:\SRETURN~x), \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \{\psi\}, \sigma)\!:\!S, H,  s, \mmap{}{\psi}{x}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

\bigskip

\begin{itemize}
\item State of the operational semantics will be $\rho, S, H, e,
  \sigma$.
\item Stack content will be $\rho, x, e, \sigma, eps$.
\item Liveness environment calculation:
  \begin{enumerate}
  \item Of the   current expression being
  evaluated:\\  
  $ \Lv = \bigcup_{ep \in \mathcal{P}(e)} \mathcal{M}_\sigma(ep)$, where $
  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$
\item Liveness environment of a stacked computation\\
  $ \Lv = \bigcup_{ep \in eps} \mathcal{M}_\sigma(ep), \text{where }
  \mathcal{M}_\sigma = \mathcal{L}(e, \sigma, \Lfonly)$ 
  \end{enumerate}
\item $GC$ models a liveness based garbage collection:
$GC(\rho, S, H, L, SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each $x \in domain(\rho)$ where $(\rho,\_,\_,\_,\_,\_)$ is a
  stack entry  with \Lv\ as the  liveness environment, $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv$.
\item for each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \notin \Lv$, .

  \end{enumerate}
\end{itemize}

>>>>

\pagebreak

\pagebreak
\section{Appendix}

We shall now  outline a proof of correctness  of the liveness analysis
presented  in  Section~\ref{sec:liveness-analysis}.   This requires  a
significant modification  to the technique  used in \cite{asati14lgc}.
\cblue{Good points to  make in the main text, right  in the small step
  semantics  section: Note  that the  state of  the  transition system
  consists of the current evaluation context, given by $\rho$ and $e$,
  and  the suspended evaluation  contexts recorded  in the  stack $S$.
  The heap $H$  is common to all contexts.}

The idea behind the proof is  as follows: We shall modify the standard
semantics to a semantics  called {\em minefield semantics} that models
a liveness based garbage collection  at each transition.  The state of
the  transition system  is augmented  to carry  enough  information to
calculate the liveness environment  at the current evaluation context,
and  each   of  the   suspended  evaluation  contexts.    Before  each
transition,  we start  from the  root-set and  insert a  special value
$\bot$ at each  location in the heap that contains  a reference and is
reachable but  not live by  our analysis.  Any attempt  to dereference
such locations during the transition will result in entering a special
state denoted \bang.  The main proof is to show that no program enters
the \bang\ state in the minefield semantics.

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$  $e,$ $\sigma,$ $e_f$.  $\sigma$ is  the demand on
  the expression and $e_f$ is the  body of the function.  The body and
  the demand are both needed to calculate the liveness environment for
  expressions.  The evaluation of  an expression may make variables in
  the scope live because of closures that are created outside it.  The
  information  in a  suspended  evaluation context  is also  similarly
  augmented with the demand on it and the function body of which the
  suspended evaluation is a part.  Thus a stacked entry now takes the
  form $(\rho, x, e, \sigma, e_f)$. A modification of the small step
  semantics which carries the extra information is shown in Figure[].
\item Given the current evaluation context $\rho, S, H, e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \sigma, e_f$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \sigma', e_f'$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \sigma, e_f$ \\ $\rightsquigarrow
  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \sigma', e_f'$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \sigma, e_f$ \\ $
  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \sigma', e_f'$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, (\clazy \cup \acar)\sigma , \_ $}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \sigma, e_f$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
    \mapsto H(\rho(x)) + H(\rho(y))], e, \sigma', e_f'$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \sigma, e_f$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \clazy\sigma, \_ $}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \sigma, e_f$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \sigma, e_g$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \sigma, e_f$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \sigma, e_f$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \sigma, e_f$ \\ $\rightsquigarrow \rho, S, H,  e_1, \sigma, e_f$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \sigma, e_f$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \sigma, e_f$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \sigma, e_f$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \sigma, e_f)\!:\!S, H, s,
\clazy\sigma, e_f$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \sigma,
  e_f$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
  \sigma', e_f'$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \sigma, e_f$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \sigma, e_f$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
